Example gene set analysis

### 2022-01-17 examples/example1.Rmd
remove mention of comb

@@

c_up <- as.data.frame(enricher(gene = de_up, universe = de_bg,  maxGSSize = 5000, TERM2GENE = genesets2))
c_up <- rownames(subset(c_up, p.adjust < 0.05))

c_dn <- as.data.frame(enricher(gene = de_dn, universe = de_bg,  maxGSSize = 5000, TERM2GENE = genesets2))
c_dn <- rownames(subset(c_dn, p.adjust < 0.05))

@@

Now performing ORA with clusterprofiler combining up and down.

@@

d_de <- as.data.frame(enricher(gene = de_de, universe = de_bg,  maxGSSize = 5000, TERM2GENE = genesets2))
d_de <- rownames(subset(d_de, p.adjust < 0.05))

@@

f_up <- as.data.frame(enricher(gene = de_up, universe = de_bg,  maxGSSize = 5000, TERM2GENE = genesets2))
f_up <- rownames(subset(f_up, p.adjust < 0.05))

f_dn <- as.data.frame(enricher(gene = de_dn, universe = de_bg, maxGSSize = 5000, TERM2GENE = genesets2))
f_dn <- rownames(subset(f_dn, p.adjust < 0.05))

@@

@@

@@

>>>>

m_up_nom <- subset(mres$enrichment_result,pANOVA<0.05 & s.dist > 0)[,1]
m_dn_nom <- subset(mres$enrichment_result,pANOVA<0.05 & s.dist < 0)[,1]

@@

c_up <- as.data.frame(enricher(gene = de_up, universe = de_bg,  maxGSSize = 5000, TERM2GENE = genesets2, pAdjustMethod="fdr"))
c_up <- rownames(subset(c_up, p.adjust < 0.05))

c_dn <- as.data.frame(enricher(gene = de_dn, universe = de_bg,  maxGSSize = 5000, TERM2GENE = genesets2, pAdjustMethod="fdr"))
c_dn <- rownames(subset(c_dn, p.adjust < 0.05))

c_up_nom <- as.data.frame(enricher(gene = de_up, universe = de_bg,  maxGSSize = 5000, TERM2GENE = genesets2, pAdjustMethod="none" ))
c_up_nom <- rownames(subset(c_up_nom, pvalue < 0.05))

c_dn_nom <- as.data.frame(enricher(gene = de_dn, universe = de_bg,  maxGSSize = 5000, TERM2GENE = genesets2, pAdjustMethod="none"))
c_dn_nom <- rownames(subset(c_dn_nom, pvalue < 0.05))

@@

Now performing ORA with clusterprofiler combining up and down. (Not shown in final article)

@@

d_de <- as.data.frame(enricher(gene = de_de, universe = de_bg,  maxGSSize = 5000, TERM2GENE = genesets2, pAdjustMethod="fdr"))
d_de <- rownames(subset(d_de, p.adjust < 0.05))

@@

f_up <- as.data.frame(enricher(gene = de_up, universe = de_bg,  maxGSSize = 5000, TERM2GENE = genesets2, pAdjustMethod="fdr"))
f_up <- rownames(subset(f_up, p.adjust < 0.05))

f_dn <- as.data.frame(enricher(gene = de_dn, universe = de_bg, maxGSSize = 5000, TERM2GENE = genesets2, pAdjustMethod="fdr"))
f_dn <- rownames(subset(f_dn, p.adjust < 0.05))

f_up_nom <- as.data.frame(enricher(gene = de_up, universe = de_bg,  maxGSSize = 5000, TERM2GENE = genesets2, pAdjustMethod="none"))
f_up_nom <- rownames(subset(f_up_nom, p.adjust < 0.05))

f_dn_nom <- as.data.frame(enricher(gene = de_dn, universe = de_bg, maxGSSize = 5000, TERM2GENE = genesets2, pAdjustMethod="none"))
f_dn_nom <- rownames(subset(f_dn_nom, p.adjust < 0.05))

@@

## Bar chart of significance

Here the idea is to classify the pathways as 1 not significant, 2 nominally significant, or 3 FDR significant

```{r,stacked1}

c_up_df <- as.data.frame(enricher(gene = de_up, universe = de_bg,  maxGSSize = 5000, TERM2GENE = genesets2, 
  pAdjustMethod="fdr" ,qvalueCutoff=1,pvalueCutoff=1))

n_c_up_ns <- nrow( subset(c_up_df,pvalue>0.05) )
n_c_up_nom <- nrow( subset(c_up_df,pvalue<0.05 ) )
n_c_up_fdr <- nrow( subset(c_up_df,p.adjust<0.05) )

c_dn_df <- as.data.frame(enricher(gene = de_dn, universe = de_bg,  maxGSSize = 5000, TERM2GENE = genesets2, 
  pAdjustMethod="fdr" ,qvalueCutoff=1,pvalueCutoff=1))

n_c_dn_ns <- nrow( subset(c_dn_df,pvalue>0.05) )
n_c_dn_nom <- nrow( subset(c_dn_df,pvalue<0.05 ) )
n_c_dn_fdr <- nrow( subset(c_dn_df,p.adjust<0.05) )

n_m_up <- length(m_up)
n_m_dn <- length(m_dn)

n_m_up_nom <- length(m_up_nom)
n_m_dn_nom <- length(m_dn_nom)

par(mar=c(5,10,5,2))

ngenes <- c("ORA up fdr"=n_c_up_fdr,"ORA up nom"=n_c_up_nom,
  "ORA dn fdr"=n_c_dn_fdr,"ORA dn nom"=n_c_dn_nom,
  "FCS up fdr"=n_m_up,"FCS up nom"=n_m_up_nom,
  "FCS dn fdr"=n_m_dn,"FCS dn nom"=n_m_dn_nom )

barplot(ngenes, horiz=TRUE,las=1)

n_c_fdr <- length(union(c_dn,c_up))
n_c_nom <- length(union(c_dn_nom,c_up_nom))
n_m_fdr <- length(subset(mres$enrichment_result,p.adjustANOVA<0.05 )[,1])
n_m_nom <- length(subset(mres$enrichment_result,pANOVA<0.05 )[,1])

par(mar=c(5,5,5,2))

ngenes <- c("ORA FDR<0.05"=n_c_fdr,"ORA p<0.05"=n_c_nom,"FCS FDR<0.05"=n_m_fdr,"FCS p<0.05"=n_m_nom)
barplot(ngenes,ylab="no. gene sets")
text((0:3*1.2)+0.7,ngenes-50,labels=ngenes,cex=1.2)

pdf("images/fdr1.pdf",width=4,height=4)
par(mar=c(5,7,2,2))
barplot(ngenes,xlab="no. gene sets",horiz=TRUE,las=1)
text( ngenes - 70, (0:3*1.2)+0.7,labels=ngenes,cex=1)
dev.off()

@@

In the next Venn I will omit combined analysis and include nominal p-value sets.

v0 <- list("FDR up"=m_up, "FDR dn"=m_dn,
           "Nom up"=m_up_nom,"Nom dn"=m_dn_nom)

plot(euler(v0),quantities = TRUE, edges = "gray", main="Effect of FDR correction on FCS results")

v0 <- list("FDR up"=c_up, "FDR dn"=c_dn,
           "Nom up"=c_up_nom,"Nom dn"=c_dn_nom)

plot(euler(v0),quantities = TRUE, edges = "gray", main="Effect of FDR correction on ORA results")

ora_nom <- union(c_up_nom,c_dn_nom)
ora_fdr <- union(c_up,c_dn)
fcs_nom <- union(m_up_nom,m_dn_nom)
fcs_fdr <- union(m_up,m_dn)

v3 <- list("ORA nom"=ora_nom, "ORA FDR"=ora_fdr,
           "FCS nom"=fcs_nom,"FCS FDR"=fcs_fdr)

plot(euler(v3),quantities = TRUE, edges = "gray", main="Effect of FDR correction")

@@

As per reviewer suggestion, remove the comparison of combined and separated ORA.

I will include using nominal p-value.

```{r,jacc2}

dat <- c("FCS vs ORA"=cm,"ORA vs ORA*"=cf,"FCS vs ORA*"=mf)

pdf("images/jacbar2.pdf",width=4,height=3)
par(mar=c(5,9,3,1))
barplot(rev(dat),xlab="Jaccard index",horiz = TRUE, las =1, xlim=c(0,.8) , main=name)
text( x=rev(dat)+0.08 , y= 1:length(rev(dat))*1.2-0.5, labels = signif(rev(dat),2))
dev.off()

png("images/jacbar2.png")
par(mar=c(5,10,3,1))
barplot(rev(dat),xlab="Jaccard index",horiz = TRUE, las =1, xlim=c(0,.75) , main=name)
text( x=rev(dat)+0.05 , y= 1:length(rev(dat))*1.2-0.5, labels = signif(rev(dat),2))
dev.off()
